---
title: "2022-12-23 Character encoding in R (windows)"
author: "Pep Porrà"
date: '2022-12-23'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

Show examples of how character encoding works in R

## Reference

This analysis is based on the document [Encoding character strings in R] (https://rstudio-pubs-static.s3.amazonaws.com/279354_f552c4c41852439f910ad620763960b6.html).
I also used additional information from [R4DS 2ed, WIP](https://r4ds.hadley.nz/strings.html)

## Steps

### Platform

```{r}
Sys.info()[1:5]
```

### System info

```{r}
sessionInfo()
```

Language is set to English in the .Rprofile file with the line `invisible(Sys.setlocale(locale = "English"))`. I revert it to my windows OS language with `Sys.setlocale()`
```{r}

```

```{r}
getOption("encoding")
```
### Localization

```{r}
l10n_info()
```

```{r}
Sys.localeconv()
```

Therefore, by default I'm using "Latin-1" encoding.

### Encoding functions

Three functions related to encoding in R base: `Encoding()`, `enc2native()` and 
`enc2utf8()`

```{r}
x_latin1 <- "Els calçots creixen millor quan 'El Niño' és molt fort"

```

```{r}
Encoding(x_latin1)
```

```{r}
x_utf8 <- enc2utf8(x_latin1)
Encoding(x_utf8)
```
### Convert to raw

```{r}
charToRaw(x_latin1)
```

```{r}
# split a string into characters
chars_in_str <- function(x){
  strsplit(x, "") |> 
    unlist()
}

# return the byte/s for each character assuming the string has the correct encoding
string_to_code <- function(y){
  sapply(chars_in_str(y), 
    function(x) paste(charToRaw(x), collapse = ":")
  )
}


```

```{r}
string_to_code(x_latin1)
```

```{r}
string_to_code(x_utf8)
```

```{r}
z <- c("–", "—")
```

```{r}
Encoding(z)
```
```{r}
string_to_code("—")
```







