---
title: "2022-12-23 Character encoding in R (windows)"
author: "Pep Porrà"
date: '2022-12-23'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(Unicode)
```


## Goal

Show examples of how character encoding works in R

## Reference

This analysis is based on:

* [Encoding character strings in R](https://rstudio-pubs-static.s3.amazonaws.com/279354_f552c4c41852439f910ad620763960b6.html)
* [R4DS (2ed), WIP](https://r4ds.hadley.nz/strings.html)
* [Encoding introduction](https://www.w3.org/International/questions/qa-what-is-encoding)

## Steps

### Platform

First, I show which platform (operating system and computer) I'm working on

```{r}
Sys.info()[1:5]
```

### System info

More extended info about my system, including the locale setup

```{r}
sessionInfo()
```

Language is set to English in my .Rprofile file with the line `invisible(Sys.setlocale(locale = "English"))`. I revert it to my windows OS language with `Sys.setlocale()`.
I've created function `locale_details()` to print locale info differently.

```{r}
locales_detail <- function(){
  info <- Sys.getlocale() |> 
    strsplit(";") |> 
    unlist() |> 
    strsplit("=") 
  v <- sapply(info, \(x) x[2])
  names(v) <- sapply(info, \(x) x[1])
  return(v)
}
```


```{r}
locales_detail()
```

```{r}
invisible(Sys.setlocale())
locales_detail()
```
By default, the global option "encoding" is set to "native.enc" 

```{r}
getOption("encoding")
```
### Localization

Function `l10n_info()` gives additional localization information. Last two 
(codepage and system.codepage) only apply to Windows systems

```{r}
l10n_info()
```

With function `Sys.localeconv()` get details of the numerical and monetary
represetantions in the current locale

```{r}
Sys.localeconv()
```

As I'm using Spanish Language, currency symbol is euro €. If locale is set to 
"English" then locale currency is USD.

```{r}
old <- Sys.setlocale(locale = "English")
Sys.localeconv()
invisible(Sys.setlocale())
```

### Encoding functions

Three functions related to encoding in R base: `Encoding()`, `enc2native()` and 
`enc2utf8()`

I'm showing how a short sentence in encode both in 'latin-1' and 'UTF-8'

```{r}
x_latin1 <- "Calçots més\n'El Niño'\n"
```

```{r}
Encoding(x_latin1)
```
```{r}
print(x_latin1)
```
```{r}
cat(x_latin1)
```

```{r}
enc2native(x_latin1) |> 
  Encoding()
```

WE can encode the sentence in 'UTF-8) with function `enc2utf8()`

```{r}
x_utf8 <- enc2utf8(x_latin1)
Encoding(x_utf8)
```
```{r}
print(x_utf8)
```
```{r}
cat(x_utf8)
```
So, enoding is completely trasparent to the user is you read a string encoded
in a given system using the same system. For strings with a defined encoding,
`print()` function uses the right encoding.

### Convert to raw

Encoding system determines the specific bytes values that represent that string.
Representation in bytes of a string is turn character into raw class. For instance,
Bytes representing the sentence used before encoded in latin-1 reads

```{r}
x_bytes <- charToRaw(x_latin1)
x_bytes
```

```{r}
class(x_bytes[1])
```
Bytes (in hexadecimal notation) can be turn into integers

```{r}
as.integer(x_bytes)
```

```{r}
as.hexmode(as.character(x_bytes[1]))
```
Function `as.hexmode()` turns strings into hexmode class objects

```{r}
as.hexmode("0xe9")
as.hexmode("e9")
print("\xe9")
print(rawToChar(as.raw("0xe9")))
```

We create a function that shows the bytes representation of each character in a
given string (that have a defined encoding)

```{r}
# split a string into characters
chars_in_str <- function(x){
  strsplit(x, "") |> 
    unlist()
}

# return the byte/s for each character assuming the string has the correct encoding
string_to_code <- function(y){
  sapply(chars_in_str(y), 
    function(x) paste(charToRaw(x), collapse = "")
  )
}


```

```{r}
string_to_code(x_latin1)
```

when encoding is "latin-1", note that each character corresponds to one byte.
Therefore, the length of the vector of characters returned by function
`string_to_code()` will be the same as the length of the string:

```{r}
nchar(x_latin1)
length(string_to_code(x_latin1))
length(charToRaw(x_latin1))
```
when using 'UTF-8' encoding, some characters are represented using two bytes

```{r}
string_to_code(x_utf8)
```

Therefore, character "é" is represented differently depending on the encoding used.

In "latin-1", "é" is "\xe9" while in "UTF-8" reads "\xc3\xa9"

```{r}
nchar(x_utf8)
length(string_to_code(x_utf8))
length(charToRaw(x_utf8))
```


```{r}
print("latin-1 Encoding")
c_latin1 <- "é"
cat("Character:", c_latin1, "\n")
cat("Encoding:", Encoding(c_latin1), "\n")
cat("Bytes representing the character:", charToRaw(c_latin1), "\n")
cat("Class of the Raw bytes:", class(charToRaw(c_latin1)), "\n") 
cat("Bytes as hexadecimals:")
print(as.hexmode(as.character(charToRaw(c_latin1))))
cat("Class of hexadecimals:", class(as.hexmode(as.character(charToRaw(c_latin1)))), "\n")
int_eq <- as.integer(charToRaw(c_latin1))
cat("Integer value:", int_eq, "\n")
cat("Unicode code for the character: ")
print(Unicode::as.u_char(int_eq))
cat("Unicode label for the character:", Unicode::u_char_label(int_eq), "\n")
```


```{r}
print("UTF-8 Encoding")
c_latin1 <- "é"
c_utf8 <- enc2utf8(c_latin1)
cat("Character:", c_utf8, "\n")
cat("Encoding:", Encoding(c_utf8), "\n")
cat("Bytes representing the character:", charToRaw(c_utf8), "\n")
cat("Class of the Raw bytes:", class(charToRaw(c_utf8)), "\n") 
cat("Bytes as hexadecimals:")
print(as.hexmode(as.character(charToRaw(c_utf8))))
cat("Class of hexadecimals:", class(as.hexmode(as.character(charToRaw(c_utf8)))), "\n")
int_eq <- utf8ToInt(c_utf8)
cat("Integer value:", int_eq, "\n")
cat("Unicode code for the character:")
print(Unicode::as.u_char(int_eq))
cat("Unicode label for the character:", Unicode::u_char_label(int_eq), "\n")
```

Unicode code can be indicated with "\\u"

```{r}
readLines(textConnection("\u00e9"))
```

```{r}
readLines(textConnection(rawToChar(as.raw("0x00E9"))))
```


```{r}
readLines(textConnection("\u00e9", encoding = "UTF-8"), encoding = "UTF-8")
```

```{r}
readLines(textConnection(x_utf8, encoding = "UTF-8"), encoding = "UTF-8") |> 
  {`[[`}(1) |> 
  string_to_code()
```

Reading a string encoded in UTF-8 in latin-1 generates issues with some characters


```{r}
readLines(textConnection(x_latin1, encoding = "UTF-8")) |> 
  {`[[`}(1) |> 
  try(string_to_code())
```

And the opposite is also true, reading in UTF-8 a text encoded in latin-1,
generates errors

```{r}
readLines(textConnection(x_latin1), encoding = "UTF-8") |> 
  {`[[`}(1) |> 
  try(string_to_code())
```

```{r}
readLines("file_latin1.txt", encoding = "latin-1")
```

```{r}
readLines("file_utf8.txt", encoding = "UTF-8")
```

If the encoding is not the right one, some characters are not read correctly

```{r}
readLines("file_latin1.txt", encoding = "UTF-8")
```

```{r}
readLines("file_utf8.txt")
```

### Print a sequence of Unicode characters

```{r}
c_ex <- c("\u00B5", "\U00BB")
```

```{r}
print(c_ex)
```

```{r}
print(c_ex)
code_c <- sapply(c_ex, utf8ToInt)
print("unicode name")
Unicode::u_char_name(code_c)
print("unicode label")
Unicode::u_char_label(code_c)
```

```{r}
class(c_ex)
```

```{r}
sapply(c_ex, charToRaw)
```

```{r}
Encoding(c_ex)
```

```{r}
sapply(c_ex, utf8ToInt)
```

Turn a UTF-8 sequence into characters

```{r}
int_series <- 1024+0:9
cat(int_series, "\n")
char_series <- intToUtf8(int_series)
cat("Bytes as hexadecimals:", "\n" )
string_to_code(char_series)
cat("word: ", char_series, "\n")
cat("# of characters: ", nchar(char_series), "\n")
cat("# of bytes: ", nchar(char_series, type = "bytes"), "\n")
cat("width: ", nchar(char_series, type = "width"), "\n")

```

### Alternative minus sign


```{r}
z <- c("+","-","–", "—", "\u2212", "\ufe63", "\uff0d")
print(z)
Encoding(z)
z_utf8 <- sapply(z, function(s) {
  if(Encoding(s) == "UTF-8") {
    return(s)
  } else {
    
    return(enc2utf8(s))
  }
})
```

```{r}
print(z_utf8)
Encoding(z_utf8)
```
```{r}
cat("UTF-8 Encoding\n---------\n")
sapply(z_utf8, string_to_code)
cat("UTF-8 into integers\n---------\n")
sapply(z_utf8, utf8ToInt)
cat("---------\nChar to raw\n")
sapply(z_utf8, charToRaw)
cat("---------\nUnicode labels\n")
sapply(z_utf8, function(x) Unicode::u_char_label(utf8ToInt(x)))
```


```{r}
cat("Latin-1 Encoding\n---------\n")
Encoding(z)
sapply(z, string_to_code)
cat("---------\nChar to raw\n")
sapply(z, charToRaw)
```
